---
title: "Bifurcations"
output: word_document
date: "2025-01-16"
---

Do logistic map and its bifurcations
Do HKB and its bifurcations

Look at critical slowing down near the bifurcation point

Can we do something in HKB to simulate behavior from the landscape?

```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "here",
                     "magrittr",
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 2 - Bifurcations.Rmd")  # To help find all the files.

install_github("barneyricca/ndstools")      # Specialized NLTS tools
library(ndstools)

source(here("Scripts/Folders.R"))           # Make sure folders exist.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

# HKB Model

Landscape and frequency changes

The HKB model has a landscape like this:
$U(\phi) = -a cos(\phi) - b cos(2 \phi)$

where $\phi$ is the relative phase between the fingers (so that $\phi = 0$ is in phase and $\phi = -\pi$ is anti-phase). (We don't care about *a* and *b*; their ratio is proportional to the frequency of flexing.)

```{r HKB}

# Go from b/a = 2 to b/a = 0 to see the bifurcation
1 -> a
.25 -> b

curve(expr = -a * cos(x) - b * cos(2*x),
      from = -4.5,
      to = 1.5,
      n = 201,
      xlab = "Phase Difference",
      ylab = "U")

```

For more details, see: Kelso, J. A. S. (1999). *Dynamic patterns: The self-organization of brain and behavior* (3rd printing). Cambridge, MA: MIT Press.

These are a different type of bifurcation than what was found in the logistic map.

