---
title: "Changepoints"
output: word_document
date: "2025-01-16"
---


```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "FNN",                 # Fast Nearest Neighbor searches
                     "here",
                     "magrittr",
                     "nonlinearTseries",
                     "tidyverse",
                     "tseries")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 8 - Changepoints.Rmd")  # To help find all the files.

install_github("barneyricca/ndstools")      # Specialized NLTS tools
library(ndstools)

source(here("Scripts/Folders.R"))           # Make sure folders exist.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default

# For testing:
#source(here("Scripts/Nonlinear cross prediction.R"))
```

# Lynx and Hare Model

We won't work through all the details here - that's what the rest of this seminar series is for - but here's the type of thing we will do. Let's ignore the fact that there are models for predator-prey populations, and work only with the data.

This run-through will largely serve as a cautionary tale for following a procedure without careful examinations along the way.

## Data
```{r}
read.csv(here("Data/Lynx and Hare.csv")) ->
  lh_df
```

## Create a Shadow State Space

Pick a stream, any stream.
```{r}
1 -> delay                                  # Arbitrary for now

plot(lh_df$Hare[-c(1:delay)],               # Omit first delay values
     lh_df$Hare[-c((nrow(lh_df)+1-delay):   # Omit last delay values
                     nrow(lh_df))],
     xlab = "h(t)",
     ylab = paste0("h(t-",
                   delay,
                   ")"),
     main = "LH Shadow State Space from Hare Population",
     type = 'l')

```

Choose a delay
```{r}
# Do NOT use this next one. You might think it works, but without additional 
#  code, it doesn't.
# mutualInformation(time.series = lh_df$Hare) ->
#   mi

mi_ksg(lh_df$Hare) ->
  mi

plot(mi)
```
The first minimum occurs at a lag of 2, so let's use that.

Create the shadow state space
```{r}
which.troughs(mi)[1] - 1 ->                 # This automates the visual
  delay                                     #  we used. 0-indexed, so - 1

```

Choose an embedding dimension. Well, there are 2 species, so I'll guess 2 (although that might turn out to be incorrect.)

Plot the shadow state space

```{r}
plot(lh_df$Hare[-c(1:delay)],               # Omit first delay values
     lh_df$Hare[-c((nrow(lh_df)+1-delay):   # Omit last delay values
                     nrow(lh_df))],
     xlab = "h(t)",
     ylab = paste0("h(t-",
                   delay-1,
                   ")"),
     main = "LH Shadow State Space from Hare Population",
     type = 'l')

```

## Tests

Noise: Is the signal simply noise?
```{r}
bds.test(x = lh_df$Hare,                    # data stream
         m = 4,                             # Embed. dim. of 2, 3, and 4
         eps = seq(0.1 * sd(lh_df$Hare),    # How close is close? Try various
                   2 * sd(lh_df$Hare),      #  values.
                   length.out = 6)) ->
  bds                                       # H0: Series of random i.i.d.
bds$p.value                                 # For various dimensions and
                                            #  definitions of "close"
```
If $p < 0.05$ there's a signal. (The time graph indicates that the signal probably isn't linear.)


Nonlinearity: Is there a nonlinear signal?
```{r}
predict_local(lh_df$Hare)$nse               # Nash-Sutcleffe coefficient of
                                            #  efficiency

```
Typically, nse > 0.65 is proposed as the threshold for a nonlinear model to exist. nse > 0 means the nonlinear model is a better predictor than the mean of the data; nse < 0 inidcates that the mean is a better predictor than the model. Of course, these are dependent on embedding dimensions, delays, and so on, so this will not always produce what is desired.

Nonstationary test: Is the signal nonstationary?
```{r}
nl_cross_pred(lh_df$Hare,
              nseg = 3,                     # Default is 5
              m.max = 6)                    # Default is 6

```

There are other tests as well, and obviously there is need to play around with parameters. In the end, however, the lynx and hare population data are very short (< 100 data points) for NLTS analyses, so one has to be very, very careful. ("Short" is typically < 1000 data points.)

## Is There a Signal?

Surrogate data testing
```{r}

```

CCM
```{r}

```

## Phenomenological Modeling

Data Pre-processing
```{r}

```


SINDy

First, SINDy on 1 data stream.
```{r}

```

Second, SINDy on the 2 data streams together.
```{r}

```

## Investigate fixed points

Fixed points and stability
```{r}

```


Does the result look anything like any of the accepted models?
