---
title: "Lorenz"
output: word_document
date: "2025-01-16"
---


```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "here",
                     "magrittr",
                     "nonlinearTseries",
                     "scatterplot3d",       # Clunky 3-d scatter plots
                     "tidyverse",
                     "tseries",             # bds.test()
                     "tseriesChaos")) {     # lorenz()
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 2 - Lorenz.Rmd")      # To help find all the files.

install_github("barneyricca/ndstools")      # Specialized NLTS tools
library(ndstools)

source(here("Scripts/Folders.R"))           # Make sure folders exist.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

# Lorenz Model

## Data

Note the paradigm here: clean signal plus noise, rather than noisy dynamics. This is just a demonstration, so we'll do this the easy way.
```{r}
lorenz() ->
  lorenz_df

set.seed(42)
0.5 ->                                      # Noise level about 1% of the data
  sigma                                     #  range.
data.frame(time = lorenz_df$time,
           x = lorenz_df$x + rnorm(length(lorenz_df$x), 0, sigma),
           y = lorenz_df$y + rnorm(length(lorenz_df$y), 0, sigma),
           z = lorenz_df$z + rnorm(length(lorenz_df$z), 0, sigma)) ->
  noisy_lorenz_df

2.5 ->                                      # Noise level about 5% of the data
  sigma                                     #  range.
data.frame(time = lorenz_df$time,
           x = lorenz_df$x + rnorm(length(lorenz_df$x), 0, sigma),
           y = lorenz_df$y + rnorm(length(lorenz_df$y), 0, sigma),
           z = lorenz_df$z + rnorm(length(lorenz_df$z), 0, sigma)) ->
  really_noisy_lorenz_df
  

# The next, and the corresponding dev.off() command, save the graph rather than
#  displaying it here.
# png(filename = here("Images/Lorenz.png"),
#     width = 10,
#     height = 5,
#     units = "in",
#     res = 600)
lorenz_df %$%
  scatterplot3d(x = x, y = y, z = z,
                type = 'l')
# dev.off()

noisy_lorenz_df %$%
  scatterplot3d(x = x, y = y, z = z,
                type = 'l')

really_noisy_lorenz_df %$%
  scatterplot3d(x = x, y = y, z = z,
                type = 'l')

```

## Create a Shadow State Space

Pick a stream, any stream...I'll go with the x-coordintate.

Choose a delay
```{r}
mi_ksg(lorenz_df$x)
```

Create the shadow state space. I'll use an embedding dimension of 3 and a delay of 14. (These are kina arbitrary.)
```{r}
embedd(lorenz_df$x,
       d = 14,
       m = 3) ->
  lorenz_shadow
```

Plot the shadow state space
```{r}
scatterplot3d(lorenz_shadow)
```

## Investigate fixed points

Fixed points and stability...trickier in 3-d...
```{r}

```

## Tests

Entropy test
```{r}
bds.test(x = lorenz_df$x,                   # data stream
         m = 4,                             # Embed. dim. of 2, 3, and 4
         eps = seq(0.1 * sd(lorenz_df$x),   # How close is close? Try various
                   2 * sd(lorenz_df$x),     #  values.
                   length.out = 6)) ->
  bds                                       # H0: Series of random i.i.d.
bds$p.value                                 # For various dimensions and
                                            #  definitions of "close"

bds.test(x = noisy_lorenz_df$x,             # data stream
         m = 4,                             # Embed. dim. of 2, 3, and 4
         eps = seq(0.1 * sd(noisy_lorenz_df$x), # How close is close? 
                   2 * sd(noisy_lorenz_df$x),   #
                   length.out = 6)) ->
  bds                                       # H0: Series of random i.i.d.
bds$p.value                                 # For various dimensions and
                                            #  definitions of "close"

bds.test(x = really_noisy_lorenz_df$x,      # data stream
         m = 4,                             # Embed. dim. of 2, 3, and 4
         eps = seq(0.1 * sd(really_noisy_lorenz_df$x), # How close is close? 
                   2 * sd(really_noisy_lorenz_df$x),   #
                   length.out = 6)) ->
  bds                                       # H0: Series of random i.i.d.
bds$p.value                                 # For various dimensions and
                                            #  definitions of "close"

```

Nonlinearity test
```{r}
#| eval: FALSE

# On my machine, the clean NSE takes more than 7 minutes, and the really noisy
#  one takes 9 minutes, so you might not want to run them now. That's why the
#  option #| eval: FALSE is at the beginning of this chunk.
# Sys.time()
# [1] "2025-02-12 11:49:50 MST"

predict_local(lorenz_df$x)$nse              # Nash-Sutcliffe efficiency
# [1] 1

# Sys.time()
# [1] "2025-02-12 11:57:02 MST"

predict_local(noisy_lorenz_df$x)$nse
# [1] 0.991

predict_local(really_noisy_lorenz_df$x)$nse
# [1] 0.844

```
Apparently, the level of noise in the system matters. We'll have to explore ways of getting rid of noise without losing the signal.

Nonstationary test
```{r}

```

## Is There a Signal?

CCM
```{r}

```

# Strange Attractors

Notice that the Lorenz butterfly exhibits a type of stability that we haven't seen before. The two "eyes" of the butterfly wings each contain a fixed-point,
and while the data never leave the vicinity of those fixed-points, they never
really approach them either. But it isn't a cyclic fixed-point, it is something new. That new thing is a *strange attractor* and those are very interesting and ubiquitous features of nonlinear systems.

# Additional Sources of Data

The Malkus Waterwheel is a physical system that is also modeled by the Lorenz system. You can find some more information about the water wheel out on the web (e.g., on Wikipedia).

You can also get lots of chaotic data by building circutis from Sprott's book,  *Elegant Circuits*.
