---
title: "Problem Demonstrations"
output: word_document
date: "2025-01-16"
---

```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # 
                     "here",
                     "magrittr",
                     "nonlinearTseries",    # Lorenz attractor
                     "phaseR",
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 1 - Problem Demonstrations.Rmd")   # To help find all the files.

source(here("Scripts/Folders.R"))           # Make sure folders exist.

install_github("barneyricca/ndstools")
library(ndstools)

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

# Short data problems
Short data. Although long-ish data streams may exist in some fields (e.g., finance), short data streams are the norm in many other fields. For example, I am currently part of a grant project that will gather 60 time points of data on each participant. This is double the number of time points that I had for other recently published projects, and 30-60 time points is not an uncommon range. Sadly, many of the methods we will use have little power at such lengths.

Khan et al., (2007) used 1000 for the cutoff for the length of "short" data series and 50-100 as the cutoff for the length of "very short" data series.

```{r prepLorenz}
lorenz(time = seq(0, 60, by = 0.01)) ->
  lorenz_ls

# The first 50 or so points are "transients," so remove them
data.frame(time = lorenz_ls$time[-c(1:100)],
           x = lorenz_ls$x[-c(1:100)],
           y = lorenz_ls$y[-c(1:100)],
           z = lorenz_ls$z[-c(1:100)]) ->
  lorenz_df

lorenz_df[1:5000,] -> lorenz_long
lorenz_df[1:500,] -> lorenz_short
lorenz_df[1:50,] -> lorenz_very_short

```

Let's use a technique from the Introduction to NDS seminar to find the fixed points

```{r lorenzFixedPoints}
fixed_1d(Level_data =                       # Doesn't work with NA
           lorenz_long$x[-nrow(lorenz_long)],
         Velocity_data = diff(lorenz_long$x)) ->
  long_fixed
{
  lorenz_long %$% plot(time, x, 
                       type = 'l',
                       main = "Long Lorenz")
  abline(h = long_fixed$Level,
         col = ifelse(long_fixed$Stability == "Stable",
                      "blue",
                      "orange"))
}

fixed_1d(Level_data =                       # Doesn't work with NA
           lorenz_short$x[-nrow(lorenz_short)],
         Velocity_data = diff(lorenz_short$x)) ->
  short_fixed
{
  lorenz_short %$% plot(time, x, 
                        type = 'l',
                        main = "Short Lorenz")
  abline(h = short_fixed$Level,
         col = ifelse(short_fixed$Stability == "Stable",
                      "blue",
                      "orange"))
}

fixed_1d(Level_data =                       # Doesn't work with NA
           lorenz_very_short$x[-nrow(lorenz_very_short)],
         Velocity_data = diff(lorenz_very_short$x)) ->
  very_short_fixed
{
  lorenz_very_short %$% plot(time, x, 
                             type = 'l',
                             main = "Very Short Lorenz")
  abline(h = very_short_fixed$Level,
         col = ifelse(very_short_fixed$Stability == "Stable",
                      "blue",
                      "orange"))
}

# long_fixed
# short_fixed
# very_short_fixed

```



# Noisy data problems
Noisy data can also be problematic. Given that nonlinear models cannot rely on noise being relatively small in relationship to the signal, any level of noise can be difficult to work with. This is exacerbated because most of the fields outside of the physical sciences and engineerings have relatively small signal-to-noise ratios.

Let's look at some examples of what can happen. First, a one-dimensional example.

```{r example1}
example1 <- function (t,                    # This is phaseR::example1
                      y, 
                      parameters) {
  list(4 - y^2)
}

findEquilibrium(deriv = example1,
                y0 = 1,
                system = "one.dim") ->
  efp1

flowField(deriv = example1,
          xlim = c(-1,1),
          ylim = c(0,5),
          system = "one.dim",
          add = FALSE)
```

Now, let's add some noise to the derivative
```{r example1_noisy}
example1_noisy <- function (t, y, parameters) {
  list(4 - y^2 + rnorm(1, 0, sd = parameters[1]))
}

1e-6 -> noise_level                         # Noise level

set.seed(42)
findEquilibrium(deriv = example1_noisy,
                y0 = 1,
                parameters = noise_level,
                system = "one.dim") ->
  efp1_noise

set.seed(42)
flowField(deriv = example1_noisy,
          xlim = c(-1,1),
          ylim = c(0,5),
          parameters = noise_level,
          system = "one.dim",
          add = FALSE)

```

Now, a 2-dimensional example:
```{r}
example15 <- function (t,                    # This is phaseR::example15
                      y, 
                      parameters) {
    list(c(y[1]^2 - 3 * y[1] * y[2] + 2 * y[1], y[1] + y[2] - 
        1))
}
findEquilibrium(deriv = example15,
                y0 = c(1,1),
                system = "two.dim") ->
  efp15

flowField(deriv = example15,
          xlim = c(-5,5),
          ylim = c(-5,5),
          system = "two.dim",
          add = FALSE)

```

```{r}
example15_noisy <- function (t,             # This is phaseR::example15
                             y, 
                             parameters) {
  list(c(y[1]^2 - 3 * y[1] * y[2] + 2 * y[1] + rnorm(1, 0, parameters[1]),
         y[1] + y[2] - 
        1 + rnorm(1, 0, parameters[1])))
}

1e-6 -> noise_level

set.seed(42)
findEquilibrium(deriv = example15_noisy,
                y0 = c(1,1),
                parameters = noise_level,
                system = "two.dim") ->
  efp15_noise

set.seed(42)
flowField(deriv = example15_noisy,
          xlim = c(-5,5),
          ylim = c(-5,5),
          parameters = noise_level,
          system = "two.dim",
          add = FALSE)

```

# Nonstationarity problems
Nonlinear dynamical systems can exhibit multiple qualitatively different behaviors. Hence, we need to be able to identify the beginning and ending of these behaviors so that our models can encompass only one behavior at a time. However, as we’ve seen, it is also possible that data which appear to be nonstationary are generated from the same dynamics that produce apparently stationary behavior.

## Phase transition

Suppose there is a sudden change in data. This occurs, for example, if a data lead on a muscle becomes detached temporarily, or there's a static build up, or a slow increase in temperature melts the solid, or whatever.

```{r}
100 -> tmax
0.01 -> beta
5.9 -> omega

1:tmax -> t
exp(-beta*t) * cos(omega*t) -> p1
0.5 + exp(-beta*t) * cos(omega*t) -> p2
-exp(-beta*t) * cos(omega*t) -> p3

c(p1, p2, p3) -> response
1:(3*tmax) -> t

plot(t, response, type = 'l')               # time-series plot
plot(response[-1], 
     response[-length(response)],
     type = 'l')

```

## Slow change

Sometimes, there is a slow change in the data's amplitude. (Slide 13 in the slide deck.)
```{r}
2000 -> N
5.3 -> num_cyc

2 * pi / N -> omega1
omega1 * num_cyc -> omega2

1:N -> t
1 -> amp

amp * (1.5 + 
         cos(omega1 * t)) *
         sin(omega2 * t) ->
  x

amp *                                       # dx is the derivative
  (1.5 + 
     cos(omega1 * t) * omega2 * cos(omega2 * 5)) -
  (sin(omega2 * t) * omega1 * sin(omega1 * t)) -> 
  dx

plot(t, x, type = 'l')                      # Plot

plot(x, dx,                                 # State space plot
     type = 'l',
     xaxt="n",
     yaxt="n")

```


# Other problems
Lastly, and this is a problem which we will grapple with more in the Part II of this series, interpreting NLTS models can be very tricky. And interpretability is important; if we cannot interpret our models, then we greatly impede the dance between theoretical and empirical studies.

Linear models are "simple" to interpret. Each regression coefficient, for example, tells us how strong the relationship is between the DV and IV.

However, any sort of interaction makes the regression no longer linear, and how do we interpret those interactions?
