---
title: "Recurrences"
output: word_document
date: "2025-01-16"
---


```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("crqa",               # R packages needed here
                     "devtools",
                     "here",
                     "magrittr",
                     "nonlinearTseries",    # Update of nonlinearAnalysis
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 4 - Recurrences.Rmd") # To help find all the files.

install_github("barneyricca/ndstools")      # Specialized NLTS tools
library(ndstools)

source(here("Scripts/Folders.R"))           # Make sure folders exist.

source(here("Scripts/plotRP.R"))            # This is the old RP from package:
                                            #  crqa. I like it better than the
                                            #  new version, although the new
                                            #  version is easier to use.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

# Converting text sequence to numeric sequence
Cleaning text is a pain. A helper function, cleanText(), is defined to make it a bit easier.

If you have rectangular numeric data (e.g., from a spreadsheet), you can skip this part.

```{r cleanText}
#| include: FALSE

# ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ###
#                                                                       #
# cleanText is a little helper function to clean up the text from the   #
#  lyrics and return a series of numbers.                               #
#  Courtesy of Aaron Likens (with some BPR modifications)               #
#                                                                       #
# x must be a single long character string that contains the text to be #
#  cleaned. I think this only works for UTF-8 encoding, though.         #
#                                                                       #
# ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ###

cleanText = function(x) {              # User-defined function
  require(tm)                          # package::tm has many text manipulation
                                       #  functions used here.
  rawText = gsub('\\n',' ',x)          # Get rid of the new line breaks
  rawText = gsub('\\r',' ',x)          # Get rid of any hard returns
  ts = Corpus(VectorSource(rawText))   # Create a time series dictionary 
                                       #  from the text.
  ts = tm_map(ts, removePunctuation)   # Yep, remove the punctuation 
  ts = tm_map(ts, removeNumbers)       # Remove numbers
  ts = tm_map(ts, tolower)             # Make everything lower case
  ts = tm_map(ts, stripWhitespace)     # Get rid of white spaces
  ts = tm_map(ts, stemDocument)        # Reduce words to their stems
  ts = tm_map(ts, PlainTextDocument)   # Make it all plain text
  ts = as.character(ts[[1]])           # Get the words as a string of numbers
  words = unlist(strsplit(ts, ' '))    # Break each number into a separate
                                       #  entry.
  return(words)                        # Send back the vector of word numbers
}
```


```{r prepText}
#| include: FALSE
#| eval: FALSE

require(readtext)                           # package:readtext is required!

# ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ###
#                                                                       #
#               USE YOUR OWN FILE IN THE NEXT COMMAND!                  #
#                                                                       #
# After that, go up to the beginning of this chunk and remove the line  #
#  that begins "#| eval: FALSE".                                        #
#                                                                       #
# ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ### ###
readtext::readtext(<your text file here>)$text -> 
  textIn

suppressWarnings(cleanText(textIn)) ->      # Clean the text
  textIn
unique(textIn) -> unique_words              # Get a list of unique words

# Derive a an ordered sequence that characterizes words in terms of their
#  order of appearance. Duplicates retain the same number:
as.vector(                                  # Need a vector, not a list
  sapply(textIn,                            # Apply to every word
         function(x) {                      # Which unique word is x?
           which(x == unique_words)})) ->   # Make these into a series
  word_series                               # word_series is a numeric vector
                                            #  suitable for RQA.

```


# RQA of Text (Songs)
Because of copyright issues, I am NOT including the data files for these songs. I have, however, put the word_series for two Beatles songs, *Let It Be* and *Revolution*, in the data folder. These data have already been cleaned and converted to numeric sequences.
```{r}
load(here("Data/Let It Be words.RData"))    # lib_words
load(here("Data/Revolution words.RData"))   # rev_words
```

If you use your own data, you must clean it according to the process in the previous section.

```{r}
rqa(time.series = lib_words,
    embedding.dim = 1,
    time.lag = 1,
    radius = 1,
    lmin = 1,
    vmin = 2,
#    save.RM = TRUE,
    do.plot = TRUE) ->
  lib_rqa

crqa(ts1 = lib_words,
     ts2 = lib_words,
     delay = 1,
     embed = 1,
     rescale = 0,
     radius = 0.05,
     normalize = 0,
     whiteline = FALSE,
     mindiagline = 2,
     minvertline = 2,
     tw = 0,
     method = "rqa",
     side = "both") ->
  lib_crqa

plot_rp(lib_crqa$RP)
```
Using the old crqa::plotRP:
```{r RQA_plot}
# set plot parameters for the recurrence plot
list(unit = 40,
     labelx = "Time", 
     labely = "Time", 
     cols = "blue", 
     pcex = .4, 
     pch = 16, 
     labax = seq(0, 
                 nrow(lib_rqa$recurrence.matrix),
                 40), 
     labay = seq(0,
                 nrow(lib_rqa$recurrence.matrix), 
                 40), 
     las = 1) -> 
  plot_params

plotRP(lib_rqa$recurrence.matrix, 
       plot_params) # generate recurrence plot
```


```{r moreData}
read.csv(here("NY Scarlet Fever.RData"))
read.csv(here("Data/econ.csv")) ->          # Interest and inflation for CRQA
  econ_df
read.csv(here("Data/Lynx and Hare.csv")) -> 
  lh_df
read.csv(here("Data/Mayport.csv")) -> 
  mp_df
read.csv(here("Data/Walking sample.csv")) -># Infant learning to walk
  walk_df

```

