---
title: "Singular Spectrum Analysis"
output: word_document
date: "2025-01-16"
---


```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "here",
                     "magrittr",
                     "Rssa",                # SSA
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Singular Spectrum Analysis.Rmd") # To help find all the files.

install_github("barneyricca/ntls_tools")    # Specialized NLTS tools
library(ntls_tools)

source(here("Scripts/Folders.R"))           # Make sure folders exist.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

```{r data}
read.csv(here("Data/Lynx and Hare")) -> lh_df
read.csv(here("Data/econ.csv")) -> econ_df
read.csv(here("Data/walking sample.csv")) -> walk_df
load(here("Data/NY Scarlet Fever.RData"))
```

Singular Spectrum Analysis

References: [Package Rssa book](https://ssa-with-r-book.github.io/) or an [introductory SSA paper](https://arxiv.org/abs/1206.6910)

From the Details of the package: "Typically the use of the package starts with the decomposition of the time series using ssa. After this a suitable grouping of the elementary time series is required. This can be done heuristically, for example, via looking at the plots of the decomposition (plot). Alternatively, one can examine the so-called w-correlation matrix (wcor). Automatic grouping can be performed by means of grouping.auto. In addition, Oblique SSA methods can be used to improve the series separability (iossa, fossa).

Next step includes the reconstruction of the time-series using the selected grouping (reconstruct)."

Let's do some

```{r SSA}
?ssa()
```


```{r SVD}
# Embedding and decomposition
ssa() ->
  ssa1

# Determine groups
plot(ssa1) # Eigenvalues 
plot(ssa1, type = "vectors")                # Eigenvectors 
plot(ssa1, type = "paired")                 # Pairs of eigenvectors 
plot(wcor(ssa1))                            # w-correlation matrix plot
```


```{r reconstructSVD}
# Reconstruct
reconstruct( 
  ssa1,
  groups = list(c(), c(), ...)) ->
  recon1
```

Toeplitz
```{r Toeplitz}

```

