---
title: "Population Models"
output: word_document
date: "2025-01-16"
---


```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "here",
                     "magrittr",
                     "tidyverse",
                     "tseries")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 2 - Population Models.Rmd")# To help find all the files.

install_github("barneyricca/ndstools")    # Specialized NLTS tools
library(ndstools)

source(here("Scripts/Folders.R"))           # Make sure folders exist.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

# Linear population model

The model:

$x_{t+1} = b x_t - d x_t = r x_t$

where *b* is the birth rate, *d* is the death rate, and *r* is the overall rate of change of population.

```{r linear_pop_gen}
lin_pop_gen <- function(x0 = 10,
                        r = 1.0,
                        num_gen = 30,
                        noise = FALSE,
                        noise_lev = 1) {

  vector(mode = "numeric",
         length = num_gen) ->
    x
  x0 -> 
    x[1]

  if(noise == TRUE) {
    for(index in 2:num_gen) {
      r * x[index - 1] + 
        rnorm(1,
              mean = 0,
              sd = noise_lev) ->
        x[index]
    }
  } else {
    for(index in 2:num_gen) {
      r * x[index - 1] ->
        x[index]
    }
  }
  return(x)
}

```

b) Show the fixed points and stabilities. r values to use: -1.2, -1, -0.5, 0, 0.5, 1, 1.2. See what types of behaviors result.

```{r lin_pop_plots}
set.seed(42)                                # For the case of adding noise

for(R in c(-1.2, -1 ,-0.5, 0, 0.5, 1, 1.2)) {
  lin_pop_gen(r = R) ->
    x
  {
    plot(1:length(x),
         x,
         xlab = "Generation",
         ylab = "Population",
         main = paste0("Growth rate: ", R))
    lines(1:length(x),
          x)
  }  
}

```

Here's another way to plot the data, a *cobweb* plot. This can be useful in helping to understand some of the transient behaviors.

```{r cobwebs}
for(R in c(-1.2, -1 ,-0.5, 0, 0.5, 1, 1.2)) {
  lin_pop_gen(r = R) ->
    x
  cobweb(x = x, title = paste("Growth rate:", R))
}
```

What about the (delay) state space? For a delay of 1 time step, that's just every other point of the cobweb plot. Or, do this type of thing

```{r LinearStateSpace}
lin_pop_gen(r = 0.5) -> x
plot(x = x[-length(x)],
     y = x[-1],
     type = 'l')
points(x = x[-length(x)],
       y = x[-1])

```

# Nonlinear Population Model

We'll look at the logistic map in a fair bit of detail, as it demonstrates most of the issues we're wanting to look at.

First, we need to be able to generate the data:

```{r logisticMap}
log_map <- function(x0 = 0.7,               # Initial value in [0,1]
                    R = 3.2,                # Growth parameter in [0,4]
                    len = 500,              # Number of generations
                    trans_len = 0,          # Length of transient
                    noise = FALSE,          # Add noise?
                    noise_lev = 0) {        # sd of added noise

  rep(0, len) ->
    x
  x0 -> x[1]
  if(noise == FALSE) {
    for(index in 2:len) {
      R * x[index-1] * (1 - x[index-1]) ->
        x[index]
    }
  } else {
    rnorm(len, 0, noise_lev) ->
      xn
    for(index in 2:len) {
      R * x[index-1] * (1 - x[index-1]) + xn[index-1] ->
        x[index]
    }
  }
  return(x)
}

```

Now, we can plot the time-series data:

```{r timePlots}
set.seed(42)

for(R in c(.5, 1.5 , 2.5, 3.1, 3.47, 3.6, 4)) {
  log_map(R = R,
          len = 150) ->
    x
  {
    plot(1:length(x),
         x,
         xlab = "Generation",
         ylab = "Population",
         main = paste0("Growth rate: ", R),
         ylim = c(0,1),                     # So all can be compared
         pch = 16,
         cex = 0.6)
    lines(1:length(x),
          x)
  }  
}

```

To note:

- Behaviors
- Transient times

We can also create cobweb plots:

```{r}
for(R in c(.5, 1.5 , 2.5, 3.1, 3.47, 3.6, 4)) {
  log_map(R = R,
          len = 150) ->
    x
  cobweb(x = x, r = R)
}

```

Back to the slides for a moment...

Let's look at fixed points. Here are the stable fixed points for the logistic map, using a *bifurcation map*. What happens to the fixed points (location and stability) as *R* varies from 0 to 4?

```{r createBifurcationData}

# Create data...lots of data...of the long-term behavior of the time-series.
#  The long-term behavior will likely be only the fixed points of the map.
2000 ->
  vec_len
1001 ->
  num_pts

list() ->
  bif_ls

seq(from = 0,
    to = 4.0,
    length.out = num_pts) ->
  R

vector(mode = "numeric",
       length = vec_len) ->
  x

for(ls_index in 1:num_pts) {
  R[ls_index] ->
    r
  0.5 -> 
    x_trans
  for(index in 1:100) {
    r * x_trans * (1 - x_trans) -> 
      xtrans
  }
  x_trans ->
    x[1]
  for(index in 2:vec_len) {
    r * x[index-1] * (1 - x[index-1]) ->
      x[index]
  }
  unique(x[-c(1:400)]) ->                   # Some transients are looooong
    bif_ls[[ls_index]]
}

```

```{r plot bifurcation}
plot(0,0,
     xlab = "R",
     ylab = "x",
     xlim = c(0,4),
     ylim = c(0,1))
for(index in 1:num_pts) {
  points(rep(R[index], length(bif_ls[[index]])),
         bif_ls[[index]],
         pch = 16,
         cex = 0.15)
}

```
What does this mean?

Incidentally, we have the fixed points for each value of R in the bif_ls[[]] list.

Let's look at the behavior of the unstable fixed points on one side of the bifurcation. Look near R = 3, and near R = 3.44949. (Really, the latter is $1 + \sqrt{6}$.) Look at time series plots and cobweb plots.
```{r}

```

Let's look at the lengths of transients near and not near bifurcation points.
We'll try R = 2.5 and R = 3.2 as being "not near" bifurcation points, and near R = 1, 3, and 3.44949 as being "near" bifurcation points. Again, look at time series plots and cobweb plots, but this time do so before the transients are removed.

```{r}

```

What tests can we do? We'll look at these in detail as the seminar moves on but here is the BDS test. This test assumes a null hypothesis that the data are generated by a random (stochastic) process. The output is a matrix of *p*-values for various *embedding dimensions* (more shortly) at different levels of assumed noise.
```{r tests}
bds.test(x)$p.value
```



