---
title: "Distance Plots"
output: word_document
date: "2025-01-16"
---


```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("crqa",               # R packages needed here
                     "devtools",
                     "here",
                     "magrittr",
                     "nonlinearTseries",    # Update of nonlinearAnalysis
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 4 - Distance Plots.Rmd") # To help find all the files.

install_github("barneyricca/ndstools")      # Specialized NLTS tools
library(ndstools)

source(here("Scripts/Folders.R"))           # Make sure folders exist.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

```{r sampleDP}

1:200 -> t
49.5 * sin(t/6) * sin(t/80) + 1 -> x

plot(t, x, 
     type = 'l', 
     lwd = 2)

outer(X = x,
      Y = x,
      FUN = "-") ->
  x_dist

colorRampPalette(c("blue",                  # Create a color palette from
                   "yellow"),                 #  blue to gold
                 alpha = TRUE) ->           # 
  this_palette
this_palette(100) ->                        # Create discrete color values 
  topo_colors

image(x = t,
      y = t,
      z = x_dist,
      zlim = range(x_dist),
      col = topo_colors,
      xaxt = 'n',
      yaxt = 'n',
      xlab = "",
      ylab = "")

```

Back to Slide Deck

# TDA

```{r}

```


# Distance Plots with TDA


Mengsen Zhang posted the data and Matlab code for Zhang et al. on [GitHub](https://github.com/mengsenz/TopoDynamics/tree/master). If you have Matlab on your computer, you can run Matlab from R, or simply go to Matlab and run Zhang's code. However, let's do the TDA stuff in R.

I copied the code, and used AI to convert it to R, then did some cleaning up (e.g., translated Matlab functions, corrected mistakes in the translation, etc.). This is what results.

```{r ZhangFigure8}
#
# This script provides an example analysis for the paper "Topological
# portraits of multiscale coordination dynamics" by Mengsen Zhang, William D.
# Kalies, J. A. Scott Kelso, and Emmanuelle Tognoli. It produces results
# shown in Fig 8 of the paper.
#
# ~~ created by Mengsen Zhang <mengsenzhang@gmail.com> 2020. ~~
#
# BPR used on online AI tool to convert Mengsen's Matlab code to R, and then 
#  modified the result to arrive at the code here. Obviously, any problems are
#  solely due to BPR and should not reflect poorly on Zhang.
#

# ===== dependencies =====
library(R.matlab)

# ===== preprocess data =====
# load either of the two files for data of examples trials: 
# "data/data_3agents.mat" or "data/data_8agents.mat"
raw_dat <- readMat(here("Data/data_8agents.mat"))

theta_original <- fullSqr2phase(raw_dat, 
                                FALSE, 
                                list('spline', NA)) 
# convert movement data to phase time series

# ===== persistent homology and recurrence plots =====
# --- parameters to extract the relevant time series
t0 <- EVENTS[3] # (s) onset of the social interaction period
tlimits <- c(1, 48) # (s) period for recurrence analysis
ds <- 25 # rate of down-sampling

# --- parameters for persistent homology
rmin <- 0.1 # smallest scale considered
rmax <- 0.5 # largest scale considered
stepsize <- 0.005 # resolution of scales
Lpnorm <- -1 # norm for compare persistent landscapes (-1=infty)

# --- regularization parameter for frequency-phase decomposition
beta <- 0.5

# --- parameters for windowing the frequency-phase graph
winsize <- 2
lag <- 1

# -- decompose original phase to a slow frequency component + a fast residual phase component
tt <- seq(tlimits[1] + t0, t0 + tlimits[2] + 1, by = winsize)
freq_theta <- thetaDecompose(theta_original, tt, sr, regularization = beta)
freq <- freq_theta[[1]]
theta_res <- freq_theta[[2]]
theta_res <- 2 * pi * theta_res

# -- segment frequency-phase graph into windows
theta_win <- windowdata(theta_res, winsize * sr, lag * sr)
freq_win <- windowdata(freq, winsize * sr, lag * sr)
t_win <- t_win / sr
X <- array(NA, dim = c(ncol(theta_win), 3, nrow(theta_win)))
X[, 1, ] <- t(theta_win)
X[, 2, ] <- t(freq_win)
X[, 3, ] <- t(t_win - matrix(t_win[, 1], nrow = nrow(t_win), ncol = ncol(t_win), byrow = TRUE))
X <- X[seq(1, dim(X)[1], ds), , ] # downsampling

# -- define distance
# distance between points in a point cloud (for topological analysis)
d_pt <- function(x, y) {
  norm(c(wrapToPi(x[1] - y[1]) / (2 * pi),
         x[2] - y[2],
         x[3] - y[3]), type = "2")
}

# distance for point-wise recurrence plot (non-topological)
d_recur <- function(x, y) {
  norm(c(wrapToPi(x[1:(length(x) / 3)] - y[1:(length(x) / 3)]) / (2 * pi),
         x[(length(x) / 3 + 1):(length(x) * 2 / 3)] - y[(length(x) / 3 + 1):(length(x) * 2 / 3)],
         x[(length(x) * 2 / 3 + 1):length(x)] - y[(length(x) * 2 / 3 + 1):length(x)]), type = "2")
}

# --- compute topological recurrence and pointwise            
result <- topoRecur(X[, , t_win[, 1] > t0], 
                    SR = 1, tlim = tlimits, distfun = d_pt, recurdist = d_recur,
                    r_max = rmax, initialstep = rmin, stepsize = stepsize, topodist = Lpnorm)
distmatpd0 <- result[[1]]
distmatpd1 <- result[[2]]
distmatrecur <- result[[3]]
pd0 <- result[[4]]
pd1 <- result[[5]]
tlim <- result[[6]]

# --- plotting the recurrence plots
par(mfrow = c(1, 3), mar = c(5, 4, 4, 2) + 0.1)
image(tlimits, tlimits, distmatpd0, axes = TRUE, xlab = 'time (s)', ylab = 'time (s)', main = 'H_0')
image(tlimits, tlimits, distmatpd1, axes = TRUE, xlab = 'time (s)', ylab = 'time (s)', main = 'H_1')
image(tlimits, tlimits, distmatrecur, axes = TRUE, xlab = 'time (s)', ylab = 'time (s)', main = 'pointwise')



```

