---
title: "Embedding"
output: word_document
date: "2025-01-16"
---


```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "here",
                     "magrittr",
                     "nonlinearTseries",    # Lorenz system
                     "scatterplot3d",
                     "tidyverse",
                     "tseriesChaos")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Embedding.Rmd")           # To help find all the files.

install_github("barneyricca/ndstools")    # Specialized NLTS tools
library(ndstools)

source(here("Scripts/Folders.R"))           # Make sure folders exist.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

# Data
```{r loadData}
read.csv(here("Data/Lynx and Hare.csv")) -> # Lynx and Hare populations
  lh_df
read.csv(here("Data/Mayport.csv")) ->       # Sea level at Mayport station
  mp_df                                     #  Has missing data (Level = -99999)
load(here("Data/NY Scarlet Fever.RData"))   # Scarlet fever in NY; messy data
```
```{r lorenzData}
suppressWarnings(                           # Deprecated behavior warning
  lorenz(sigma = 10,                        # Traditional parameters are 10,
         beta = 8/3,                        #  8/3, and 28
         rho = 28,
         start = c(-13, -14, 47),           # Will converge to butterfly
         time = seq(0, 50, by = 0.01))) ->  # 5001 data points
  lorenz_df
```


# Constant Delay Embedding
- Choose delay (AMI)
- Choose Theiler window (ACF)
- Choose embedding dimension (FNN)

```{r autoEmbed}

```
```{r embedByHand}

```

# Principal Component Analysis Embedding
- Lots of components
- Which contribute a lot?
```{r PCAExample}
# If everyone is familiar with PCA, we'll skip this. Otherwise, we'll do it.

```
```{r PCAEmbedding}
# Start by embedding the Lorenz data in a much higher dimension than you
#  think you need. Takens can help here, but that comes later
11 ->                                       # embedding dimension
  m
1 ->                                        # time delay; choose a small value
  d
embedd(lorenz_df$x,                         # Need the TseriesChaos version
      m,                                    #  to override the default auto-
      d) ->                                 #  behavior.
  emb

pc<- prcomp(emb,                            # Principal Components Analysis
            center = TRUE,                  # Center and scale the variables.
            scale. = TRUE)                  #
summary(pc)  

plot(100 * pc$sdev / sum(pc$sdev ^ 2),      # Scree plot
     type="b",
     pch = 19,
     cex = 1.2,
     xlab="Principal Component", 
     ylab="Percent Variance",
     ylim=c(0,40),  
     xlim=c(1,m),
     cex.lab=1.5,
     cex.axis=1.2)  

```
```{r plotPCAReconstruction}
scatterplot3d(predict(pc)[,1:3],
              type="l",
              cex.lab=1.5,
              cex.axis=1.2,
              lab=c(4,4,7),
              lab.z=4)
```

# Maximally Orthogonal Embedding
- Construct delay for each additional dimension
- Stop when no additional information
- Does better compared to the "flat" dimensions that can show up in the constant delay embedding.
```{r maxOrthoEmbed}
max_ortho_embed(lorenz_df$x) ->
  emb_mo

emb_mo$dims
emb_mo$delays

scatterplot3d(emb_mo$ss[,1:3],
              type="l",
              cex.lab=1.5,
              cex.axis=1.2,
              lab=c(4,4,7),
              lab.z=4)

```

# Multi-Dimension Embedding


